---
title: "R-functions"
author: "Francesco Pupillo"
date: "02/10/2020"
output: 
  #github_document:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R functions

This is a collection of functions typically used in R, especially in psychology.  

For any question, feel free to drop me an email : pupillo@psych.uni-frankfurt.de

### Exclude rows based on a set of criteria (or participants)
Useful code to exclude rows in a dataset that corresponds to participants we decided to exclude. 

```{r exclusionSeveralCriteria, echo=TRUE}

# create fake dataframe with participants and performance columns
df<-data.frame(participants=seq(1:10), performance= sample(seq(0, 1, length.out = 100), 10, T))

df
# Participants that we want to exclude
partExcl<-c(1,3,5)

# code to exclude participants
df<-df[!df$participants %in% partExcl, ]

df

# we could also use dplyr
library(dplyr)

df<- df %>%
  filter(!participants %in% partExcl)

```
### Count the arguments of a function
Function that returns the number of arguments in a function
```{r count the arguments of a function, echo=TRUE}
# let's create a function that calculate the means between three dimenions (x, y ,z)

foo<-function(x, y, z){
  sum (x, y, z)/ length(c(x, y,z))
}

# now I want to assign the number of the arguments of this function to a vector. 
# I can use "formals" to store the argumens

fooArgs<-formals(foo)

fooArgs

# then I can extract the number of the arguments with "length"
numfooArgs<-length(fooArgs)

numfooArgs
```

### Call a function by a character string
How to assign a name of a function to a character string and then call it later? By using "get".
```{r call a function by a character string, echo=TRUE}
# Create a function. This one calculate the average of three dimensions

foo<-function(x, y, z){
  sum (x, y, z)/ length(c(x, y,z))
}

# Now create a carachter string with the name of the formula

foostr<-"foo"

foostr

# use "get" to call the function

myfunction<-get(foostr)

# now we can call our function
myfunction(3,4,5)

numfooArgs
```




### Go through a list of items and select the ones that meet some criteria (contains rm in a loop)
This example is based on the BOSS dataset
``` {r select items}
# we need the readxl package to read the excel file with the object info
library(readxl)
# and dplyr for intabulating
library(dplyr)

# retrieve the file with info of the categories
category<-read_excel("SI1.xlsx", sheet = "Sheet2")

# change fourth and fifth names to include an underscore
names(category)[4:5]<-c("modal_categ", "cat_agreement")

# get the number of objects per category
table<-category %>% 
  group_by(modal_categ) %>% 
  tally()

# order by modal categ and name agreement
table<-table[order(table$n, decreasing=T),]

# show the first 11
head(table, n=11)

# select the categories
selCat<-c( "Outdoor activity & sport item", "Kitchen & utensil","Electronic device & accessory", "Hand labour tool & accessory")

# create a dataset with only the categories
dataSel<-category[category$modal_categ %in% selCat, ]

# select the the first 40 images with the highest category agreement level for each category, 
# assign them to the relative dataset
for (cat in 1:length(selCat)){
  # subset the database
  dataim<- dataSel[dataSel$modal_categ==selCat[cat],]
  # order it
  dataim<-dataim[order(dataim$cat_agreement, decreasing = T),]
  # select first 40 
  datapractsel<-dataim[1:40,]
  # assign to a dataset
  assign(paste(selCat[cat],"_dat",sep=""), datapractsel)
  
}

# merget the datasets 
tomerge<-ls(pattern = "*_dat$")
dataim_all<-as.data.frame(NULL)

for (n in tomerge){
  dataim_all<-rbind(dataim_all, get(n) )
}

# delete the previously created single datasets
for (i in 1:length(tomerge)){
  name<-tomerge[i]
  rm(list=tomerge[i])
}

```
### select items in a dataset that start or end with a string
```{R, eval = F}
# select variables that start with lh or rh
datasub<-data[, c( grep( "^lh_", names(data), value = TRUE), 
                   grep( "^rh_", names(data), value = TRUE))]

# select items that ends with "thickness"
datasub<-data[, c(grep("_thickness$", names(data), value = TRUE))]

```



### Select elements in a vector that start or end with a letter
```{r}

# create a vector
vect<-c("One", "Two", "Three", "Four", "Five")

# select only the elements that start with the letter "T"
Tvect<-vect[grep("T", vect)]

Tvect

# select only the elements that end with the letter "e"
Evect<-vect[grep("e$", vect)]

Evect
```
### Looping
Functions that are used for looping through data
#### create a progress bar
```{r progress bar}

# using the previous example of the boss dataset

# make a progress bar
pb<-txtProgressBar(min=0, max=length(selCat), style =3)

for (cat in 1:length(selCat)){
  # subset the database
  dataim<- dataSel[dataSel$modal_categ==selCat[cat],]
  # order it
  dataim<-dataim[order(dataim$cat_agreement, decreasing = T),]
  # select first 40 
  datapractsel<-dataim[1:40,]
  # assign to a dataset
  assign(paste(selCat[cat],"_dat",sep=""), datapractsel)
  
  #progress bar
  setTxtProgressBar(pb, i) 
}

```

#### Append dataframes in a loop
```{r eval=FALSE, include=T}
# first, create a list
my_list<-list()


# loop through the dataframes
for (dataframe in 1:length(dataframes)){
  
  currdataframe<-dataframes[datamframe]
  # assign a dataframe to the list
  my_list[[dataframe]]<-currdataframe
  
  
}

# merge the files outside the loop
all_df<-do.call(rbind, my_list)


```


#### Error Handling
Don't stop the loop when there is an error, but catch the error
```{r error handling, eval=F}
# create a variable with random numbers, but with a NA
randVar<-c(1,2,3,4,NA, 6,7)

# loop through it, checking whether it is bigger than 5
for (num in randVar){
  if (num>5){
    print(paste(num, "is bigger than five"))
  } else {
    print(paste(num, "is smaller than five"))
  }
}

# create a variable to store missing cases
missing_cases<-vector()

# and a counter to store them
miss_count<-1
# as you can see, it stops when there is a NA
# let's use 'tryCatch'
for (num in randVar){
  tryCatch({
    
    if (num>5){
      print(paste(num, "is bigger than five"))
    } else {
      print(paste(num, "is smaller than five"))
    }
  },
  
  # we want to cathc the error and print something informative,
  # like at which number it crashed
  error= function(e) {print(paste("problem with number", num))
    
    print(paste(num, "is not a number"))
    
    # we could also do something for this case
    # we could assign to a list of missing cases, so we can count them
    missing_cases[miss_count]<<-num
    # we need the superassignment operator "<<-" to assign this to the 
    # global environment
    
    # update the counter
    miss_count<<-miss_count+1
  }
  )
}

# how many missing cases?
length(missing_cases)

```

### Dplyr
#### Count number of observations by group
```{r}
category %>%
  group_by(modal_categ) %>%
  tally()
```
#### Count number of missing cases by group

```{r}
category %>%
  group_by(modal_categ) %>%
  dplyr::summarise(CountCase = n(), PercNA = sum(is.na(modal_categ))/n()*100)
```


#### Calculate means with grouping factor with dplyr
```{r}
library(dplyr)
meanGroup<- category %>%
  group_by(modal_categ) %>%
  slice( (1:10)) %>% # take only the first 10
  summarise(meanCateg = mean(cat_agreement),
            Dataset = unique(Dataset)) # keep the other variable

meanGroup


```

#### Add trial number in a long dataset with dplyr
```{r}

# create data frame 
trial_df <- data.frame('subject' = c(rep('101', 3), rep('102', 3)),
                       'result' = c(32, 33, 64, 12, 18, 14))

# group by subject and add a row number
# assumes your data frame is ordered by trial number for each subject
trial_df <- trial_df %>%
  group_by(subject) %>%
  mutate(trial_number = row_number())
```

### ggplot

#### Plot a summary line across the trials
```{r warning=FALSE}
# load the ggplot library
library(ggplot2)

# import the toy dataset
data_long<-read.csv("All_data.csv")

sum_et<-data_long %>%
  group_by(participant) %>%
  summarise(et = mean(fixation_error, na.rm = T))

# which participants have the et data?
part_incl<-sum_et$participant[!is.na(sum_et$et)]
# select only those
all_data_et<-data_long[data_long$participant %in% part_incl,]

# reorder the levels
all_data_et$age_group<-factor(all_data_et$age_group, levels = c("CH", "YA", "OA"))
all_data_et$type<-factor(all_data_et$type, c("low", "medium", "high", "singletons"))

# look at the data
head(all_data_et[c("participant", "fixation_prediction", "age_group", "trial_n")])

# now we want to plot a line which summarises the fixation-deviation across the trials, with some errors

# we need to summarise the percentage within participants
library("Rmisc") # load the library

# sumamrise
dat_summary_fix_pr<- summarySEwithin(all_data_et,
                                     measurevar = "fixation_prediction",
                                     withinvars = "trial_n" , 
                                     betweenvars = "age_group",
                                     idvar = "participant", 
                                     na.rm = T)

# conver the trial number as numeric
dat_summary_fix_pr$trial_n<-as.numeric(dat_summary_fix_pr$trial_n)
# load the library
library(ggplot2)

# plot
ggplot(dat_summary_fix_pr,aes(x = trial_n, y = fixation_prediction, 
                              # this assigns the color to the group
                              color = age_group, fill = age_group, 
                              # this changes the line type as a funciton of group
                              linetype = age_group, group = 1))+
  # this adds the line
  stat_summary(fun.y="mean",geom="line", size = 1.5)+
  # this add the shadow considering the within-participant standard error
  geom_ribbon(aes(ymin=fixation_prediction-se, ymax=fixation_prediction+se), alpha=0.5, colour=NA)+
  # divide the plot as a funtion of age group
  facet_wrap(.~age_group)+
  # customise the breaks
  scale_x_continuous(breaks=seq(1, 100, 19))+
  
  # add personalized colours
   scale_color_manual(values = c(c( "#AA4499" ,"#44AA99","#332288")))+

   # add personalized parmaeters
    theme(
    plot.title = element_text(size = 30),
    axis.title.x = element_text(size = 28),
    axis.title.y = element_text(size = 28),
    axis.text=element_text(size=28),
    legend.text=element_text(size=rel(2)),
    legend.title = element_text(size=rel(2)), 
    strip.text.x = element_text(size=28)
  )+
  theme(plot.title = element_text(hjust = 0.5))+
  # use the classic theme
  theme_classic()
    # add annotation


```

#### Set default options for multiple plots 
```{r warning=FALSE}
# We need to set a function
custom_param<-function(){

       # add personalized parmaeters
    theme(
    plot.title = element_text(size = 30),
    axis.title.x = element_text(size = 28),
    axis.title.y = element_text(size = 28),
    axis.text=element_text(size=28),
    legend.text=element_text(size=rel(2)),
    legend.title = element_text(size=rel(2)), 
    strip.text.x = element_text(size=28)
   )+
  theme(plot.title = element_text(hjust = 0.5))
}

# now plot with those parameters
ggplot(dat_summary_fix_pr,aes(x = trial_n, y = fixation_prediction, 
                              # this assigns the color to the group
                              color = age_group, fill = age_group, 
                              # this changes the line type as a funciton of group
                              linetype = age_group, group = 1))+
  # this adds the line
  stat_summary(fun.y="mean",geom="line", size = 1.5)+
  custom_param()+
  facet_wrap(.~age_group)+
  theme_classic()

```


#### Plot within-particiants error bars plus individual lines by condition
```{r warning=FALSE}
# first, summarise with within-participant error bars
dat_summary_fix_error<- summarySEwithin(all_data_et,
                                       measurevar = "fixation_error",
                                       withinvars = c("type" ), 
                                       betweenvars = "age_group",
                                       idvar = "participant", 
                                       na.rm = T)

# plot it
# first, aggregate the data by participant, age group, and condition (type)
ggplot(all_data_et %>%
         group_by(participant,age_group, type)%>%
         dplyr::summarise(fixation_error=mean(fixation_error, na.rm=T)), 
       aes(x = type, y = fixation_error, colour = type ))+
  # Add some points, which correspond to the participants
  geom_point(alpha = 0.10, colour = "black" )+
  # add a line that connect those points by participant
  geom_line( aes(type, fixation_error,group = participant),
             size=1, alpha=0.1, stat="summary" , colour = 'black')+
  # add a summary line (mean)
  geom_point(stat="summary", size = 5, data = dat_summary_fix_error)+
  xlab("")+
  # add the within-participant confidence intervals
  geom_errorbar(aes( y = fixation_error, ymin = fixation_error - ci, 
                     ymax = fixation_error + ci),
                width = 0.40,size = 1.5, data=dat_summary_fix_error)+
  # divide the plot as a function of age group
  facet_wrap(.~age_group)+
  theme_classic()+
  custom_param()+
  # delete the x axis text and ticks, as we have the legend
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank()) 

```
#### Spaghetti plot
```{r}
# Create a spaghetti plot of some subset trials
  ggplot(all_data_et[all_data_et$trial_n<21,], 
         aes( x=trial_n, y=fixation_prediction))+
  # add the "smooth" line, which the regression method ('l,')
  # and trasparent (0.5)
    geom_line(stat="smooth",method = "lm", formula=y~x, alpha=0.5, se=F)+

  # specify that we want different colours for different participants
    aes(colour = factor(participant))+
  # add the summary line with geom_smooth
    geom_smooth(method="lm",formula=y~x, se=T, colour = "black" )+
    theme(strip.text.x = element_text(size = 13))+
    theme_classic()+
    theme(panel.spacing = unit(1, "lines"))+
    facet_wrap(.~age_group)+
    #ggtitle("Experiment 2")+
    theme(legend.position = "none")


```
#### Spaghetti plot with quadratic
```{r}
all_data_et$PE<-as.numeric(all_data_et$PE)
ggplot(all_data_et, aes( x=PE, y=conf_resp.keys))+
  geom_line(stat="smooth",method = "lm", formula=y~poly(x,2), alpha=0.5, se=F)+
  aes(colour = factor(participant))+
  geom_smooth(method="lm",formula=y~poly(x,2), se=T, colour = "black" )+
  theme(strip.text.x = element_text(size = 13))+
  theme_classic()+
  theme(panel.spacing = unit(1, "lines"))+
  facet_wrap(.~age_group)+
  #ggtitle("Experiment 2")+
  theme(legend.position = "none")
```

#### Density plots 
```{r}
# Density plots
Plot_loc<-ggplot(all_data_et, aes(x= location_error, fill=type))

Plot_loc+
# plot the density in trasparency
  geom_density(alpha = .5)+
  theme_classic()+
  facet_wrap(.~age_group)+
  
  # geom_vline(xintercept = thres)+
  theme(
    plot.title = element_text(size = 22),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    axis.text=element_text(size=20)
  )+
  xlab("location error")+
  #theme(legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))

```


### System Administration Tasks

#### Creating, deleting, copying files and directories
``` {r create delete copy folders, eval=F}
# create a directory
dir.create("new_folder")

# create a file
# file.create("new_text_file.txt")
# file.create("new_word_file.docx")
# file.create("new_csv_file.csv")

# assign a matrix to those files
fileSample<-matrix(data =0, nrow = 2, ncol = 2)
write.table(fileSample,"new_text_file.txt", row.names = F )
write.table(fileSample,"new_text_file.docx", row.names = F )
write.csv(fileSample,"new_csv_file.csv", row.names = F )


# create lots of files
sapply(paste0("new_folder/file", 1:100, ".txt"), file.create)

# copy files
file.copy("source_file.txt", "destination_folder")

# list all CSV files non-recursively
list.files(pattern = ".csv")

# list all CSV files recursively through each sub-folder
list.files(pattern = ".csv", recursive = TRUE)

# read in all the CSV files
all_data_frames <- lapply(list.files(pattern = ".csv"), read.csv)

# stack all data frames together
single_data_frame <- Reduce(rbind, all_data_frames)

# remove files
file.remove("new_text_file.txt")

# check if a file exists
file.exists("new_text_file.txt")

# check if a folder exists
file.exists("new_folder")
```

#### Get the script's directory path
Get the path to directory that contains the script and set it as the working directory 
```{r script directory, echo=TRUE}

# get the script directory
path<-rstudioapi::getSourceEditorContext()$path

# split the string into the names of the folders
names<-unlist(strsplit(path, split="/"))

# get number of carachters last name (file name)
charfile<-nchar(tail(names,1))

# subtract that to the path
path<-substr(path, 1,nchar(path) - charfile)

# set wd 
setwd(path)
```

#### Select files according to characters that come before a symbol
Return the characters before a specific symbol
```{r symbol, echo=TRUE}

# take the first file from the previous example
setwd("testsel")
files<-list.files( pattern= ".csv$")
name<-files[1]

name
# Select the characters before the ".pdf"
sel<-sub("\\.pdf.*", "", name)

sel
```

#### Select files that meet a criterion
Selecting files that meets a criterion, or file estension
```{r select files, echo=TRUE}
# create a vector with names that vary in different ways
files<-vector()

for (i in 1:10){
  # the first five are .csv, the others are .pdf
  if (i<6){
    files[i]<- paste("myfile", i, ".csv", sep="")
  } else{
    files[i]<- paste("myfile", i, ".pdf", sep="")
  }
}

# plus some random files
files[11:13]<-c("random1.pdf", "test2.csv", "dunno.rnd")

# create the files in a subfolder
# dir.create("testsel")

setwd("testsel")

# create files
for (f in 1:length(files)){
  write(0, file = files[f])
}



# select all the files that are .csv
selCsv<-list.files( pattern= ".csv$")
selCsv

# now, select files that are .csv OR that start with "myfile"
selCsv<-list.files(pattern= c("myfile", ".csv$"))
selCsv

# now, select files that are .csv AND that start with "myfile"
selCsv<-list.files(pattern= c("myfile.*.csv$"))
selCsv

```
## Stats
### Sumamry SE within with non-normative means
SummarySEwithin from the "Rmisc" package is a very handy function to compute
within-participant errors (ci, sd, and se). H
However, the mean returned by this function is normative. If you use summary
SE within together with ggplot, you might want to use the following version, 
which return the non-normative means.

Taken from:

```{r}
summarySE2 <- function (data = NULL, measurevar, groupvars = NULL, na.rm = TRUE, conf.interval = 0.95) {
  library(data.table)
  data <- data.table(data)
  
  length2 <- function(x, na.rm = FALSE) {
    if (na.rm) 
      sum(!is.na(x))
    else length(x)
  }
  
  datac <- data[, .(unlist(lapply(.SD, length2, na.rm = na.rm)), 
                    unlist(lapply(.SD, mean, na.rm = na.rm)),
                    unlist(lapply(.SD, sd, na.rm = na.rm))),
                by = groupvars, .SDcols = measurevar]
  names(datac) <- c(groupvars, "N", measurevar, "sd")
  setkeyv(datac, groupvars)
  
  datac[, se := unlist(sd) / sqrt(unlist(N))] #compute standard error
  
  ciMult <- qt(conf.interval / 2 + 0.5, unlist(datac$N) - 1)
  datac[, ci := se * ciMult]
  datac <- data.frame(datac)
  return(datac)
}

normDataWithin2 <- function (data = NULL, idvar, measurevar, betweenvars = NULL, 
                             na.rm = TRUE) {
  library(data.table); library(dplyr)
  data <- data.table(data)
  setkeyv(data, idvar)
  
  data.subjMean <- data[, .(unlist(lapply(.SD, mean, na.rm = na.rm))), by = idvar, .SDcols = measurevar]
  names(data.subjMean) <- c(idvar, 'subjMean')
  data <- merge(data, data.subjMean)
  setkeyv(data, c(idvar, betweenvars))
  
  measureNormedVar <- paste(measurevar, "Normed", sep = "")
  data <- data.frame(data)
  
  data[, measureNormedVar] <- data[, measurevar] - unlist(data[, "subjMean"]) + mean(data[, measurevar], na.rm = na.rm)
  return(data)
}


#normed and un-normed versions
summarySEwithin2 <- function (data = NULL, measurevar, betweenvars = NULL, withinvars = NULL, 
                              idvar = NULL, na.rm = TRUE, conf.interval = 0.95) {
  
  # Ensure that the betweenvars and withinvars are factors
  factorvars <- sapply(data[, c(betweenvars, withinvars), drop = FALSE], 
                       FUN = is.factor)
  if (!all(factorvars)) {
    nonfactorvars <- names(factorvars)[!factorvars]
    message("Automatically converting the following non-factors to factors: ", 
            paste(nonfactorvars, collapse = ", "))
    data[nonfactorvars] <- lapply(data[nonfactorvars], factor)
  }
  
  # Get the means from the un-normed data
  datac <- summarySE2(data, measurevar, groupvars=c(betweenvars, withinvars),
                      na.rm=na.rm, conf.interval=conf.interval)
  
  # Drop all the unused columns (these will be calculated with normed data)
  datac$sd <- NULL
  datac$se <- NULL
  datac$ci <- NULL
  
  # Norm each subject's data
  ndata <- normDataWithin2(data, idvar, measurevar, betweenvars, na.rm)
  
  # This is the name of the new column
  measurevar_n <- paste(measurevar, "Normed", sep="")
  
  # Collapse the normed data - now we can treat between and within vars the same
  ndatac <- summarySE2(ndata, measurevar_n, groupvars=c(betweenvars, withinvars),
                       na.rm=na.rm, conf.interval=conf.interval)
  
  # Apply correction from Morey (2008) to the standard error and confidence interval
  #  Get the product of the number of conditions of within-S variables
  nWithinGroups    <- prod(vapply(ndatac[,withinvars, drop=FALSE], FUN= function(x) length(levels(x)),
                                  FUN.VALUE=numeric(1)))
  correctionFactor <- sqrt( nWithinGroups / (nWithinGroups-1) )
  
  # Apply the correction factor
  ndatac$sd <- unlist(ndatac$sd) * correctionFactor
  ndatac$se <- unlist(ndatac$se) * correctionFactor
  ndatac$ci <- unlist(ndatac$ci) * correctionFactor
  
  # Combine the un-normed means with the normed results
  merged <- merge(datac, ndatac)
  #merged[, 1] <- as.numeric(as.character(merged[, 1]))
  #merged <- merged[order(merged[, 1]), ]
  return(merged)
}



```


