---
title: "R-functions"
author: "Francesco Pupillo"
date: "02/10/2020"
output: 
  #github_document:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R functions

This is a collection of functions typically used in R, especially in psychology.  

For any question, feel free to drop me an email : pupillo@psych.uni-frankfurt.de

### Exclude rows based on a set of criteria (or participants)
Useful code to exclude rows in a dataset that corresponds to participants we decided to exclude. 

```{r exclusionSeveralCriteria, echo=TRUE}

# create fake dataframe with participants and performance columns
df<-data.frame(participants=seq(1:10), performance= sample(seq(0, 1, length.out = 100), 10, T))

df
# Participants that we want to exclude
partExcl<-c(1,3,5)

# code to exclude participants
df<-df[!df$participants %in% partExcl, ]

df

# we could also use dplyr
library(dplyr)

df<- df %>%
  filter(!participants %in% partExcl)

```
### Count the arguments of a function
Function that returns the number of arguments in a function
```{r count the arguments of a function, echo=TRUE}
# let's create a function that calculate the means between three dimenions (x, y ,z)

foo<-function(x, y, z){
  sum (x, y, z)/ length(c(x, y,z))
}

# now I want to assign the number of the arguments of this function to a vector. 
# I can use "formals" to store the argumens

fooArgs<-formals(foo)

fooArgs

# then I can extract the number of the arguments with "length"
numfooArgs<-length(fooArgs)

numfooArgs
```

### Call a function by a character string
How to assign a name of a function to a character string and then call it later? By using "get".
```{r call a function by a character string, echo=TRUE}
# Create a function. This one calculate the average of three dimensions

foo<-function(x, y, z){
  sum (x, y, z)/ length(c(x, y,z))
}

# Now create a carachter string with the name of the formula

foostr<-"foo"

foostr

# use "get" to call the function

myfunction<-get(foostr)

# now we can call our function
myfunction(3,4,5)

numfooArgs
```




### Go through a list of items and select the ones that meet some criteria (contains rm in a loop)
This example is based on the BOSS dataset
``` {r select items}
# we need the readxl package to read the excel file with the object info
library(readxl)
# and dplyr for intabulating
library(dplyr)

# retrieve the file with info of the categories
category<-read_excel("SI1.xlsx", sheet = "Sheet2")

# change fourth and fifth names to include an underscore
names(category)[4:5]<-c("modal_categ", "cat_agreement")

# get the number of objects per category
table<-category %>% 
  group_by(modal_categ) %>% 
  tally()

# order by modal categ and name agreement
table<-table[order(table$n, decreasing=T),]

# show the first 11
head(table, n=11)

# select the categories
selCat<-c( "Outdoor activity & sport item", "Kitchen & utensil","Electronic device & accessory", "Hand labour tool & accessory")

# create a dataset with only the categories
dataSel<-category[category$modal_categ %in% selCat, ]

# select the the first 40 images with the highest category agreement level for each category, 
# assign them to the relative dataset
for (cat in 1:length(selCat)){
  # subset the database
  dataim<- dataSel[dataSel$modal_categ==selCat[cat],]
  # order it
  dataim<-dataim[order(dataim$cat_agreement, decreasing = T),]
  # select first 40 
  datapractsel<-dataim[1:40,]
  # assign to a dataset
  assign(paste(selCat[cat],"_dat",sep=""), datapractsel)
  
}

# merget the datasets 
tomerge<-ls(pattern = "*_dat")
dataim_all<-as.data.frame(NULL)

for (n in tomerge){
  dataim_all<-rbind(dataim_all, get(n) )
}

# delete the previously created single datasets
for (i in 1:length(tomerge)){
  name<-tomerge[i]
  rm(list=tomerge[i])
}

```
### select items in a dataset that start or end with a string
```{R, eval = F}
# select variables that start with lh or rh
 datasub<-data[, c( grep( "^lh_", names(data), value = TRUE), 
             grep( "^rh_", names(data), value = TRUE))]

# select items that ends with "thickness"
datasub<-data[, c(grep("_thickness$", names(data), value = TRUE))]

```



### Select elements in a vector that start or end with a letter
```{r}

# create a vector
vect<-c("One", "Two", "Three", "Four", "Five")

# select only the elements that start with the letter "T"
Tvect<-vect[grep("T", vect)]

Tvect

# select only the elements that end with the letter "e"
Evect<-vect[grep("e$", vect)]

Evect
```
### Looping
Functions that are used for looping through data
#### create a progress bar
```{r progress bar}

# using the previous example of the boss dataset

# make a progress bar
pb<-txtProgressBar(min=0, max=length(selCat), style =3)

for (cat in 1:length(selCat)){
  # subset the database
  dataim<- dataSel[dataSel$modal_categ==selCat[cat],]
  # order it
  dataim<-dataim[order(dataim$cat_agreement, decreasing = T),]
  # select first 40 
  datapractsel<-dataim[1:40,]
  # assign to a dataset
  assign(paste(selCat[cat],"_dat",sep=""), datapractsel)
  
  #progress bar
  setTxtProgressBar(pb, i) 
}

```

#### Append dataframes in a loop
```{r eval=FALSE, include=T}
# first, create a list
my_list<-list()


# loop through the dataframes
for (dataframe in 1:length(dataframes)){
  
  currdataframe<-dataframes[datamframe]
  # assign a dataframe to the list
  my_list[[dataframe]]<-currdataframe
  
  
}

# merge the files outside the loop
all_df<-do.call(rbind, my_list)


```


#### Error Handling
Don't stop the loop when there is an error, but catch the error
```{r error handling, eval=F}
# create a variable with random numbers, but with a NA
randVar<-c(1,2,3,4,NA, 6,7)

# loop through it, checking whether it is bigger than 5
for (num in randVar){
  if (num>5){
    print(paste(num, "is bigger than five"))
  } else {
    print(paste(num, "is smaller than five"))
  }
}

# create a variable to store missing cases
missing_cases<-vector()

# and a counter to store them
miss_count<-1
# as you can see, it stops when there is a NA
# let's use 'tryCatch'
for (num in randVar){
  tryCatch({
    
    if (num>5){
      print(paste(num, "is bigger than five"))
    } else {
      print(paste(num, "is smaller than five"))
    }
  },
  
  # we want to cathc the error and print something informative,
  # like at which number it crashed
  error= function(e) {print(paste("problem with number", num))
    
    print(paste(num, "is not a number"))
    
    # we could also do something for this case
    # we could assign to a list of missing cases, so we can count them
    missing_cases[miss_count]<<-num
    # we need the superassignment operator "<<-" to assign this to the 
    # global environment
    
    # update the counter
    miss_count<<-miss_count+1
  }
  )
}

# how many missing cases?
length(missing_cases)

```

### Dplyr
#### Calculate means with grouping factor with dplyr
```{r}
library(dplyr)
meanGroup<- category %>%
  group_by(modal_categ) %>%
  slice( (1:10)) %>% # take only the first 10
  summarise(meanCateg = mean(cat_agreement),
            Dataset = unique(Dataset)) # keep the other variable

meanGroup


```

#### Add trial number in a long dataset with dplyr
```{r}

# create data frame 
trial_df <- data.frame('subject' = c(rep('101', 3), rep('102', 3)),
                  'result' = c(32, 33, 64, 12, 18, 14))

# group by subject and add a row number
# assumes your data frame is ordered by trial number for each subject
trial_df <- trial_df %>%
  group_by(subject) %>%
  mutate(trial_number = row_number())
```



### System Administration Tasks

#### Creating, deleting, copying files and directories
``` {r create delete copy folders, eval=F}
# create a directory
dir.create("new_folder")

# create a file
# file.create("new_text_file.txt")
# file.create("new_word_file.docx")
# file.create("new_csv_file.csv")

# assign a matrix to those files
fileSample<-matrix(data =0, nrow = 2, ncol = 2)
write.table(fileSample,"new_text_file.txt", row.names = F )
write.table(fileSample,"new_text_file.docx", row.names = F )
write.csv(fileSample,"new_csv_file.csv", row.names = F )


# create lots of files
sapply(paste0("new_folder/file", 1:100, ".txt"), file.create)

# copy files
file.copy("source_file.txt", "destination_folder")

# list all CSV files non-recursively
list.files(pattern = ".csv")

# list all CSV files recursively through each sub-folder
list.files(pattern = ".csv", recursive = TRUE)

# read in all the CSV files
all_data_frames <- lapply(list.files(pattern = ".csv"), read.csv)

# stack all data frames together
single_data_frame <- Reduce(rbind, all_data_frames)

# remove files
file.remove("new_text_file.txt")

# check if a file exists
file.exists("new_text_file.txt")

# check if a folder exists
file.exists("new_folder")
```

#### Get the script's directory path
Get the path to directory that contains the script and set it as the working directory 
```{r script directory, echo=TRUE}

# get the script directory
path<-rstudioapi::getSourceEditorContext()$path

# split the string into the names of the folders
names<-unlist(strsplit(path, split="/"))

# get number of carachters last name (file name)
charfile<-nchar(tail(names,1))

# subtract that to the path
path<-substr(path, 1,nchar(path) - charfile)

# set wd 
setwd(path)
```

#### Select files according to characters that come before a symbol
Return the characters before a specific symbol
```{r symbol, echo=TRUE}

# take the first file from the previous example
setwd("testsel")
files<-list.files( pattern= ".csv$")
name<-files[1]

name
# Select the characters before the ".pdf"
sel<-sub("\\.pdf.*", "", name)

sel
```

#### Select files that meet a criterion
Selecting files that meets a criterion, or file estension
```{r select files, echo=TRUE}
# create a vector with names that vary in different ways
files<-vector()

for (i in 1:10){
  # the first five are .csv, the others are .pdf
  if (i<6){
    files[i]<- paste("myfile", i, ".csv", sep="")
  } else{
    files[i]<- paste("myfile", i, ".pdf", sep="")
  }
}

# plus some random files
files[11:13]<-c("random1.pdf", "test2.csv", "dunno.rnd")

# create the files in a subfolder
# dir.create("testsel")

setwd("testsel")

# create files
for (f in 1:length(files)){
  write(0, file = files[f])
}



# select all the files that are .csv
selCsv<-list.files( pattern= ".csv$")
selCsv

# now, select files that are .csv OR that start with "myfile"
selCsv<-list.files(pattern= c("myfile", ".csv$"))
selCsv

# now, select files that are .csv AND that start with "myfile"
selCsv<-list.files(pattern= c("myfile.*.csv$"))
selCsv

```